---
title: "TBR 2026 — Race Strategy"
date: "2026-04-03"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    embed-resources: true
    fig-width: 12
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(xml2)
library(knitr)
library(kableExtra)

# ── Parameters (must match daily_distances.R) ────────────────────────────────
GPX_FILE        <- "Trans Balkan Race 2026 __ Issue 1.GPX"
SLEEP_HOUR      <- 2
SLEEP_DURATION  <- 4.5
MAX_SLEEP       <- 6
MIN_RIDING_H    <- 10
MAX_DAILY_CLIMB <- 4000
BASE_SPEED      <- 13.8
MAX_DESCENT     <- 32
UPHILL_K        <- 0.08
DOWNHILL_K      <- 0.04
FTP             <- 237  # watts

race_start <- as.POSIXct("2026-05-29 09:00:00", tz = "Europe/Paris")

# ── Parse GPX & build segments ───────────────────────────────────────────────
gpx  <- read_xml(GPX_FILE); ns <- c(g = "http://www.topografix.com/GPX/1/1")
pts  <- xml_find_all(gpx, ".//g:trkpt", ns)
track <- tibble(
  lat = as.numeric(xml_attr(pts, "lat")),
  lon = as.numeric(xml_attr(pts, "lon")),
  ele = as.numeric(xml_text(xml_find_first(pts, "g:ele", ns)))
) |>
  mutate(ele = zoo::rollmedian(ele, k = 31, fill = NA, align = "center")) |>
  filter(!is.na(ele))

haversine <- function(lat1, lon1, lat2, lon2) {
  r <- 6371; phi1 <- lat1*pi/180; phi2 <- lat2*pi/180
  a <- sin(((lat2-lat1)*pi/180)/2)^2 + cos(phi1)*cos(phi2)*sin(((lon2-lon1)*pi/180)/2)^2
  2 * r * asin(sqrt(a))
}

segments <- track |>
  mutate(dist_km = haversine(lag(lat), lag(lon), lat, lon),
         ele_diff = ele - lag(ele),
         grade = pmax(pmin(ifelse(dist_km > 0, (ele_diff/(dist_km*1000))*100, 0), 30), -30)) |>
  slice(-1) |> filter(!is.na(dist_km), dist_km > 0) |>
  mutate(
    speed_kmh = case_when(
      grade >= 0 ~ BASE_SPEED * exp(-UPHILL_K * grade),
      grade <  0 ~ pmin(BASE_SPEED * exp(-DOWNHILL_K * grade), MAX_DESCENT)),
    time_h = dist_km / speed_kmh, gain_m = pmax(ele_diff, 0),
    cum_dist = cumsum(dist_km), cum_time = cumsum(time_h), cum_gain = cumsum(gain_m))

total_dist  <- sum(segments$dist_km)
total_climb <- sum(segments$gain_m)
total_time  <- sum(segments$time_h)

# ── Day simulation ───────────────────────────────────────────────────────────
simulate_days <- function(segs, start_wall) {
  WAKE_HOUR <- SLEEP_HOUR + SLEEP_DURATION
  days_list <- list(); seg_idx <- 1L; wall_time <- start_wall
  while (seg_idx <= nrow(segs)) {
    day_h0 <- as.numeric(difftime(wall_time, floor_date(wall_time, "day"), units = "hours"))
    hrs_to_latest <- (SLEEP_HOUR - day_h0 + 24) %% 24
    hrs_to_earliest <- (WAKE_HOUR - MAX_SLEEP - day_h0 + 24) %% 24
    if (hrs_to_latest == 0) hrs_to_latest <- 24
    if (hrs_to_earliest == 0) hrs_to_earliest <- 24
    day_riding <- 0; day_climb <- 0; day_dist <- 0; day_start <- wall_time
    while (seg_idx <= nrow(segs)) {
      s <- segs[seg_idx, ]
      if (day_riding + s$time_h > hrs_to_latest) break
      if (day_climb + s$gain_m > MAX_DAILY_CLIMB &&
          day_riding >= MIN_RIDING_H && day_riding >= hrs_to_earliest) break
      day_riding <- day_riding + s$time_h; day_dist <- day_dist + s$dist_km
      day_climb <- day_climb + s$gain_m; seg_idx <- seg_idx + 1L
    }
    stop_wall <- day_start + as.difftime(day_riding, units = "hours")
    next_0630 <- floor_date(stop_wall, "day") + as.difftime(WAKE_HOUR, units = "hours")
    if (next_0630 <= stop_wall) next_0630 <- next_0630 + days(1)
    sleep_h <- max(SLEEP_DURATION, min(MAX_SLEEP,
                 as.numeric(difftime(next_0630, stop_wall, units = "hours"))))
    days_list[[length(days_list) + 1]] <- list(
      day_start_wall = day_start, day_end_wall = stop_wall,
      riding_h = day_riding, dist_km = day_dist, climb_m = day_climb, sleep_h = sleep_h)
    wall_time <- stop_wall + as.difftime(sleep_h, units = "hours")
    if (seg_idx > nrow(segs)) break
  }
  bind_rows(lapply(days_list, as_tibble)) |>
    mutate(day = row_number(), cum_dist = cumsum(dist_km), cum_riding = cumsum(riding_h))
}

day_info <- simulate_days(segments, race_start)
finish_wall <- day_info$day_start_wall[nrow(day_info)] +
  as.difftime(total_time - c(0, day_info$cum_riding)[nrow(day_info)], units = "hours")

# ── Helpers ──────────────────────────────────────────────────────────────────
riding_time_at_dist <- function(dist_km)
  approx(segments$cum_dist, segments$cum_time, xout = dist_km, rule = 2)$y

wall_clock <- function(riding_time_h) {
  day_idx <- pmin(findInterval(riding_time_h, day_info$cum_riding) + 1L, nrow(day_info))
  cum_start <- c(0, day_info$cum_riding)[day_idx]
  day_info$day_start_wall[day_idx] + as.difftime(riding_time_h - cum_start, units = "hours")
}

# ── Checkpoints & resupply ───────────────────────────────────────────────────
checkpoints <- tribble(
  ~name,           ~cum_dist_km, ~cutoff,
  "CP1 Mazin",            451,   as.POSIXct("2026-06-01 17:00:00", tz = "Europe/Paris"),
  "CP2 Kopilovi",        1060,   as.POSIXct("2026-06-04 23:59:00", tz = "Europe/Paris"),
  "Finish Risan",        1395,   as.POSIXct("2026-06-07 23:59:00", tz = "Europe/Paris")
) |>
  mutate(
    riding_h = map_dbl(cum_dist_km, riding_time_at_dist),
    est_arrival = do.call(c, map(riding_h, wall_clock)),
    margin_h = as.numeric(difftime(cutoff, est_arrival, units = "hours"))
  )

resupply <- tribble(
  ~location,              ~km,   ~type,         ~on_route,
  "Kozina (SLO)",         16.6,  "Village",      TRUE,
  "Slavnik (SLO)",        28.2,  "Mountain hut", TRUE,
  "Dane (CRO)",           47.4,  "Fountain",     TRUE,
  "Sveti Pavel (SLO)",    74.2,  "Fountain",     TRUE,
  "Platak (CRO)",        141,    "Mountain hut", TRUE,
  "Fužine (CRO)",        164,    "Village",      TRUE,
  "Oltari (CRO)",        245,    "Fountain",     TRUE,
  "Dom Zavižan (CRO)",   260,    "Mountain hut", TRUE,
  "Gospić (CRO)",        345,    "City",         TRUE,
  "CP1 Mazin (CRO)",     451,    "Checkpoint",   TRUE,
  "Velika Popina (CRO)", 490,    "Fountain",     TRUE,
  "Knin (CRO)",          537,    "City",         TRUE,
  "Livno (BiH)",         664,    "Village",      TRUE,
  "Šuica (BiH)",         701,    "Village",      TRUE,
  "Mostar (BiH)",        812,    "City",         TRUE,
  "Nevesinje (BiH)",     858,    "Village",      FALSE,
  "Ulog (BiH)",          897,    "Village",      TRUE,
  "Fountain (BiH)",      938,    "Fountain",     TRUE,
  "Miljevina (BiH)",     990,    "Village",      FALSE,
  "Popov Most (BiH)",   1034,    "Village",      TRUE,
  "Brod (BiH)",         1055,    "Village",      TRUE,
  "CP2 Kopilovi (BiH)", 1060,    "Checkpoint",   TRUE,
  "Žabljak (MNE)",      1160,    "City",         TRUE,
  "Njegovuđa (MNE)",    1170,    "Village",      TRUE,
  "Kolašin (MNE)",      1228,    "City",         FALSE,
  "Nikšić (MNE)",       1325,    "City",         TRUE,
  "Risan / Finish",     1395,    "Finish",       TRUE
) |>
  mutate(
    riding_h = map_dbl(km, riding_time_at_dist),
    day_num  = findInterval(riding_h, day_info$cum_riding) + 1L,
    arrival  = do.call(c, map(riding_h, wall_clock)),
    gap_km   = km - lag(km, default = 0),
    gap_h    = riding_h - lag(riding_h, default = 0)
  )
```

## The numbers

```{r overview-table}
tibble(
  ` ` = c("Distance", "Climbing", "Riding time (model)", "Sleep schedule",
          "Estimated finish", "Target", "FTP"),
  `  ` = c(
    sprintf("%.0f km", total_dist),
    sprintf("%.0f m", total_climb),
    sprintf("%.1f h", total_time),
    "02:00-06:30 (4.5h min, 6h max)",
    format(finish_wall, "%a %d %b %H:%M CEST"),
    "< 9 days (finish by Sat 6 Jun)",
    sprintf("%d W", FTP)
  )
) |> kable(col.names = NULL)
```

::: {.callout-important}
## Model vs reality
The simulation assumes 100% riding efficiency (no stops). Your Hydra Epic data shows **65-72% efficiency** on multi-day stages (stops for eating, navigation, mechanicals, bathroom). A 19.5h riding window realistically yields **13-14h of pedaling**. The model's 7-day finish is optimistic; **7.5-8.5 days** is the realistic target at your pace.
:::

## Checkpoint timeline

```{r checkpoint-table}
checkpoints |>
  mutate(
    `Est. arrival` = format(est_arrival, "%a %d %b %H:%M"),
    Cutoff = format(cutoff, "%a %d %b %H:%M"),
    `Margin` = sprintf("%.0fh", margin_h)
  ) |>
  select(Checkpoint = name, km = cum_dist_km, `Est. arrival`, Cutoff, Margin) |>
  kable() |>
  kable_styling(full_width = FALSE)
```

::: {.callout-note}
Even with realistic efficiency losses, margins are comfortable. CP1 has **30+ hours** of buffer. The biggest risk is not cutoffs but accumulated fatigue on Days 5-7.
:::

## Day-by-day strategy

```{r day-data}
# Build detailed day data
day_detail <- day_info |>
  mutate(
    start_dist = lag(cum_dist, default = 0),
    date_str = format(day_start_wall, "%a %d %b"),
    window = sprintf("%s–%s", format(day_start_wall, "%H:%M"), format(day_end_wall, "%H:%M")),
    # Get elevation stats per day
    max_elev = map2_dbl(start_dist, cum_dist, ~max(segments$ele[segments$cum_dist > .x & segments$cum_dist <= .y], na.rm = TRUE)),
    min_elev = map2_dbl(start_dist, cum_dist, ~min(segments$ele[segments$cum_dist > .x & segments$cum_dist <= .y], na.rm = TRUE))
  )
```

### Day 1 — Friday 29 May: Trieste to ~km 179

**Slovenia and Croatian coast → mountains**

```{r day1-stats}
d <- day_detail[1,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~11h (first day conservative)",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            "First 50km: 1530m climb. Steep coastal mountains.")
) |> kable(col.names = NULL)
```

**Strategy:**

- **Resist the start adrenaline.** Day 1 sets the tone for 8 more days. You know this from Hydra E1.
- The opening 50 km is brutal: **1530m of climbing** at 5.9% average grade. This is the Slovenian karst. Don't try to keep up with anyone.
- **Power cap: 130-140W average** (55-59% FTP). HR stays under 135.
- At Hydra E1 you averaged 132W and HR 118 over 25.4h — that's the benchmark.
- **Nutrition from the first hour**: 60-80g carbs/hr. Don't wait until hungry.
- First resupply: Kozina (km 17), Slavnik mountain hut (km 28). Stock up in Fužine (km 164, village with shops).
- **Sleep target: Stop by 00:30**, sleep until 06:30 (6h sleep on night 1 is worth it).

::: {.callout-warning}
## Critical gap: Sveti Pavel → Platak (67 km, 1533m climb, 5.8h)
No resupply for 67 km through mountain terrain. Carry food and water from Sveti Pavel (km 74).
:::

### Day 2 — Saturday 30 May: ~km 179 to ~km 409

**Croatian highlands — Velebit mountains**

```{r day2-stats}
d <- day_detail[2,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~14h",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            "Velebit range. Steep grades (avg 9.2% on climbs). Gravel/forest roads.")
) |> kable(col.names = NULL)
```

**Strategy:**

- Wake 06:30. Coffee, eat 500 kcal breakfast. Rolling by 07:00.
- This is the **steepest average climbing day** (9.2% on uphills). Seated climbing, low cadence, Z2-Z3.
- **Power target: 120-130W average**. You'll push 160-200W on climbs but average will drop on rough terrain.
- Key resupply: Oltari (km 245), Dom Zavižan mountain hut (km 260), Gospić (km 345, city — **big resupply**).
- The Velebit section (km 200-300) is remote: **1200m+ climbing per 50km**, poor surfaces.

::: {.callout-warning}
## Critical gap: Fužine → Oltari (81 km, 1667m climb, 7.2h)
Remote mountain section. Carry 3000+ kcal and 2+ liters.
:::

### Day 3 — Sunday 31 May: ~km 409 to ~km 646

**Lika plateau → Dalmatian hinterland**

```{r day3-stats}
d <- day_detail[3,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "CP arrival", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~14h",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            sprintf("CP1 Mazin (~km 451) around %s", format(checkpoints$est_arrival[1], "%H:%M")),
            "Mix of gravel and paved. Enters Bosnia late in day.")
) |> kable(col.names = NULL)
```

**Strategy:**

- **CP1 Mazin at ~km 451** — early in the day. Cutoff is Mon 1 Jun 17:00; you'll arrive **~30h ahead**.
- After CP1: Gospić to Knin section. Resupply at Knin (km 537, city).
- Fatigue is building. Stick to the routine: eat every 30 min, drink every 15 min.
- **Power will naturally drop to 110-120W average** — this is normal. Don't chase Day 1 numbers.

::: {.callout-warning}
## Critical gap: Knin → Livno (127 km, 2058m climb, 10.5h)
**The biggest gap on the entire route.** No resupply for 127 km through remote terrain. Load up fully in Knin: 5000+ kcal, 3+ liters. This gap spans from late Day 3 into Day 4.
:::

### Day 4 — Monday 1 June: ~km 646 to ~km 882

**Central Bosnia — mountains and valleys**

```{r day4-stats}
d <- day_detail[4,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~14h",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            "Bosnian mountains. Gravel, some paved through towns. Mostar at km 812.")
) |> kable(col.names = NULL)
```

**Strategy:**

- This is the **mental halfway point**. 4 days done, 4 to go. Expect a psychological dip.
- Mostar (km 812) is the biggest city since Trieste — **major resupply opportunity**. Get a proper meal, fill everything.
- Power target: 110-120W. Accept it. Hydra showed you drop 20% by mid-race.
- **Watch for appetite crash** around Day 3-4. Force yourself to eat even if food sounds terrible. Switch to savory/salty foods if sweet becomes unbearable.

::: {.callout-tip}
## Mostar (km 812)
Real city. Restaurants, shops, pharmacies. If you need bike parts, this is the place. Consider a 30-45min longer stop for a hot meal and mental reset.
:::

### Day 5 — Tuesday 2 June: ~km 882 to ~km 1091

**Eastern Bosnia — the hard middle**

```{r day5-stats}
d <- day_detail[5,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~13h",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            "Remote mountain terrain. Hardest climbing section (km 900-950: 1349m/50km).")
) |> kable(col.names = NULL)
```

**Strategy:**

- **4102m of climbing** — second hardest day. The section km 900-950 (after Ulog) is particularly steep.
- Approaching CP2 Kopilovi (km 1060). Arrival estimated around evening.
- **This is the day that breaks people.** Fatigue is maximal, terrain is hard, remoteness is real.
- Keep the routine. Wake, eat, ride, eat, ride, eat, sleep. Nothing else matters.
- Watch for the fountain at km 938 — it may be the only water for a long stretch.

::: {.callout-warning}
## Critical gap: Ulog → Fountain (41 km, 1116m climb, 3.9h)
Followed by Fountain → Miljevina (52 km, 747m climb, 4.2h). Remote Bosnian mountains. Carry extra water.
:::

### Day 6 — Wednesday 3 June: ~km 1091 to ~km 1290

**Montenegro mountains — the hardest day**

```{r day6-stats}
d <- day_detail[6,]
tibble(
  Metric = c("Window", "Riding (model)", "Realistic riding", "Distance", "Climbing",
             "Elevation range", "CP arrival", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h), "~13h",
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            sprintf("CP2 Kopilovi (~km 1060) early if not reached Day 5"),
            "HARDEST DAY. 4619m climbing. Durmitor and Montenegro mountains. Avg 9.9% on climbs.")
) |> kable(col.names = NULL)
```

**Strategy:**

- **4619m of climbing** — the biggest day. Max elevation 1880m. Average uphill grade **9.9%**.
- Section km 1200-1250 is the steepest on the whole route: **1387m in 50km**.
- CP2 Kopilovi if not reached on Day 5 — cutoff is Thu 4 Jun 23:59, you have huge margin.
- **Think of Hydra E7-E8**: you rode 147km/1696m on Day 7 and 330km/3156m on Day 8. You've been deeper than this.
- Žabljak (km 1160) is a proper city in Montenegro — resupply here.
- Power: accept 100-115W average. Your Hydra E7-E8 power was 100-105W. This is normal.

::: {.callout-warning}
## Critical gap: CP2 Kopilovi → Žabljak (100 km, 2335m climb, 8.9h)
And Kolašin → Nikšić (97 km, 1963m climb, 8.5h). Both are massive gaps in Montenegro mountains.
:::

### Day 7 — Thursday 4 June: ~km 1290 to Finish (km 1395)

**The final 97 km — descent to Risan**

```{r day7-stats}
d <- day_detail[7,]
tibble(
  Metric = c("Window", "Riding (model)", "Distance", "Climbing",
             "Elevation range", "Key terrain"),
  Value = c(d$window, sprintf("%.1fh", d$riding_h),
            sprintf("%.0f km", d$dist_km), sprintf("%.0f m", d$climb_m),
            sprintf("%.0f–%.0f m", d$min_elev, d$max_elev),
            "Only 715m climbing. Mostly descent from mountains to Kotor Bay (sea level).")
) |> kable(col.names = NULL)
```

**Strategy:**

- **The easy day.** Only 97 km and 715m of climbing. Mostly downhill to Kotor Bay.
- Don't crash on the descents. Tired riders crash. **Brake before corners**, not in them.
- Nikšić (km 1325) is the last city before the finish — last resupply if needed.
- **Enjoy the final descent to Risan.** You've earned it.

::: {.callout-tip}
Model estimates arrival around **13:30 on Thu 4 June** — 6.2 days. With realistic efficiency, expect **finish Fri 5 Jun or Sat 6 Jun**, well within your 9-day target and days before the cutoff.
:::

## Power strategy

```{r power-zones}
tibble(
  Context = c("Flat / tailwind", "Rolling terrain", "Climbing (< 6%)",
              "Steep climbing (> 8%)", "Descending", "Race average target"),
  `Power (W)` = c(
    sprintf("130-165 (55-70%% FTP)", FTP),
    sprintf("120-150", FTP),
    sprintf("165-200 (70-84%% FTP)", FTP),
    sprintf("140-180 (60-76%% FTP)", FTP),
    "Soft pedal / coast",
    "100-130"
  ),
  `HR cap` = c("<135", "<135", "<140", "<145", "N/A",
               "<130"),
  Notes = c(
    "Sustainable all day, every day",
    "Natural variation is fine",
    "Don't surge — steady effort up",
    "Sit and spin, don't stand unless necessary",
    "Rest the legs, eat, drink",
    "Your Hydra avg was 105-132W across 8 stages"
  )
) |> kable() |> kable_styling(full_width = FALSE)
```

### Power decay over multi-day

Your Hydra Epic data shows a clear power decay pattern:

```{r power-decay}
# Hydra data from analysis
hydra_stages <- tribble(
  ~stage, ~day, ~km, ~hours, ~avg_watts, ~avg_hr,
  "E1", 1, 429, 25.4, 132, 118,
  "E2", 2, 202, 14.1, 109, 107,
  "E3", 3, 183, 12.1, 115, 109,
  "E4", 4, 199, 12.5,  80, 107,
  "E5", 5, 205, 12.9, 114, 106,
  "E6", 6, 155, 11.7,  98, 102,
  "E7", 7, 147, 11.5, 100, 102,
  "E8", 8, 330, 23.1, 105, 102
)

ggplot(hydra_stages, aes(day, avg_watts)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 3) +
  geom_hline(yintercept = 132, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 1.5, y = 136, label = "Day 1: 132W", size = 3, colour = "grey40") +
  geom_hline(yintercept = 105, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 7, y = 109, label = "Day 7-8: 100-105W", size = 3, colour = "grey40") +
  scale_x_continuous(breaks = 1:8) +
  labs(title = "Hydra Epic 2025 — Average Power by Stage",
       subtitle = "Expect ~20% power drop from Day 1 to Day 7+",
       x = "Race day", y = "Average power (W)") +
  theme_minimal(base_size = 12)
```

**Implications for TBR:**

- Day 1: ~125-130W avg → realistic
- Days 2-3: ~115-120W avg → settling in
- Days 4-5: ~110-115W avg → fatigue accumulating
- Days 6-7: ~100-110W avg → deep fatigue, this is expected
- **Don't fight the decay.** It's physiology. Maintain effort, accept lower power.

## Nutrition plan

### Hourly targets

```{r nutrition-table}
tibble(
  Riding = c("Hours 1-4", "Hours 5-10", "Hours 10-14", "At stops"),
  `Carbs/hr` = c("80g", "70-80g", "60-70g", "Eat a full meal if >15min stop"),
  `Fluid/hr` = c("500-750ml", "500-750ml", "500ml", "Drink to thirst"),
  `What to eat` = c(
    "Gels, bars — whatever is convenient",
    "Mix gels/bars with real food (sandwich, banana)",
    "Savory foods, rice, bread — sweet becomes repulsive",
    "Whatever hot food is available. Soup, pasta, pizza."
  )
) |> kable() |> kable_styling(full_width = FALSE)
```

### Daily nutrition targets

| Metric | Target | Notes |
|---|---|---|
| Total kcal intake | 6000-8000 | You'll burn 8000-10000, but gut limits absorption |
| Carbohydrates | 500-700g | Main fuel source |
| Protein | 100-150g | From real food at resupply stops |
| Sodium | 2000-3000mg | More in heat; salt tablets or electrolyte mix |
| Caffeine | Reserve for nights | Cola at gas stations, caffeine gels after midnight |

### Resupply strategy

```{r resupply-critical}
resupply |>
  mutate(
    arrival_fmt = format(arrival, "%a %d %b %H:%M"),
    location = if_else(!on_route, paste0(location, " *"), location)
  ) |>
  select(Location = location, km, `+km` = gap_km, Type = type,
         Day = day_num, `Est. arrival` = arrival_fmt) |>
  mutate(km = sprintf("%.0f", km), `+km` = sprintf("%.0f", `+km`)) |>
  kable(align = c("l","r","r","l","c","l")) |>
  kable_styling(full_width = FALSE, font_size = 12) |>
  row_spec(which(resupply$type == "Checkpoint"), bold = TRUE, background = "#fff3cd") |>
  row_spec(nrow(resupply), bold = TRUE, background = "#d4edda") |>
  row_spec(which(resupply$gap_km > 80), color = "red")
```

`*` = off-route

### Critical resupply gaps (> 80 km)

```{r critical-gaps}
resupply |>
  filter(gap_km > 60) |>
  arrange(desc(gap_km)) |>
  mutate(
    section_climb = map2_dbl(km - gap_km, km, function(from, to) {
      sum(segments$gain_m[segments$cum_dist > from & segments$cum_dist <= to])
    })
  ) |>
  select(
    `From → To` = location, `Gap km` = gap_km, `Climb` = section_climb,
    `Riding hours` = gap_h, Day = day_num
  ) |>
  mutate(`Gap km` = sprintf("%.0f", `Gap km`),
         Climb = sprintf("%.0fm", Climb),
         `Riding hours` = sprintf("%.1fh", `Riding hours`)) |>
  kable() |>
  kable_styling(full_width = FALSE) |>
  row_spec(1:3, bold = TRUE, color = "red")
```

::: {.callout-important}
## The 3 biggest gaps
1. **Knin → Livno (127 km)**: Fill everything in Knin. Carry 5000+ kcal, 3L water.
2. **Šuica → Mostar (111 km)**: Stock up in Šuica.
3. **Gospić → CP1 Mazin (106 km)**: Last big resupply in Gospić (city with shops).
:::

## Sleep strategy

### Schedule

| | Target | Notes |
|---|---|---|
| **Stop riding** | 00:30-02:00 | Hard stop. No riding after 02:00. |
| **Sleep** | 02:00-06:30 (4.5h) | Minimum. Take 6h on Night 1 and after hardest days. |
| **Wake** | 06:30 | Same time every day. Routine = survival. |
| **Rolling** | By 07:00 | 30min for wake-up, eat, pack, bathroom |

### Sleep execution

1. **Scout bivvy spot 15-20 min before stopping.** Don't ride until 02:00 then spend 30min looking.
2. Target: **15 min from stopping to sleeping.** Practice this before the race.
3. Setup: bivvy bag/tent → sleeping bag → earplugs → eye mask → alarm set → sleep.
4. **Don't skip sleep to make time.** Sleep deprivation costs more than it saves after Day 3.

### Night riding

- After Day 2, you'll be riding in darkness from ~21:00-00:30 every night.
- **Lights**: Helmet + bar mount. Spare batteries. USB charging from dynamo or power bank.
- **Navigation**: GPS visibility is critical at night. Mount at eye level.
- **Drowsiness**: If you nod off on the bike, stop immediately. 15-20min "power nap" is better than crashing.

## Efficiency analysis

Your Hydra Epic efficiency data:

```{r efficiency}
hydra_eff <- tribble(
  ~stage, ~moving_h, ~elapsed_h,
  "E1", 25.4, 37.8,
  "E2", 14.1, 20.3,
  "E3", 12.1, 17.1,
  "E4", 12.5, 17.4,
  "E5", 12.9, 18.2,
  "E6", 11.7, 18.1,
  "E7", 11.5, 16.9,
  "E8", 23.1, 37.1
) |>
  mutate(efficiency = moving_h / elapsed_h * 100,
         stop_h = elapsed_h - moving_h)

hydra_eff |>
  select(Stage = stage, `Moving (h)` = moving_h, `Elapsed (h)` = elapsed_h,
         `Stops (h)` = stop_h, `Efficiency (%)` = efficiency) |>
  mutate(across(where(is.numeric), ~sprintf("%.1f", .))) |>
  kable() |> kable_styling(full_width = FALSE)
```

**Key insight:** You spend **5-6 hours per day NOT riding** (eating, navigation, mechanicals, bathroom, transitions). For TBR planning:

- **19.5h available window** (06:30-02:00) × 70% efficiency = **~13.5h actual riding**
- At your speed model, that's **140-170 km/day** — right on target for sub-9-day finish
- To improve: reduce stop time (eat on the bike, fast bivvy setup, less dithering at junctions)

## Equipment checklist

### Critical (race-ending if missing)

| Category | Items |
|---|---|
| **Navigation** | GPS with full TBR route loaded, backup phone with offline maps, USB charger |
| **Lights** | Helmet light + bar light, spare batteries, USB charging cable |
| **Repair** | Spare tube ×2, patch kit, tire boot, multi-tool, chain quick link, tire lever |
| **Sleep** | Bivvy bag or ultralight tent, sleeping bag/quilt (comfort 5-10°C), inflatable pillow |
| **Safety** | Emergency bivvy (reflective), phone with local emergency numbers, ID + insurance card |

### Important (race-impacting if missing)

| Category | Items |
|---|---|
| **Clothing** | Rain jacket (waterproof), arm warmers, leg warmers, wind vest, spare chamois |
| **Nutrition** | Frame bag for food, 2+ bottle cages, electrolyte mix, caffeine gels for night |
| **Comfort** | Chamois cream, anti-chafing, lip balm with SPF, sunscreen |
| **Electronics** | Power bank (10000+ mAh), charging cables, earplugs, eye mask |

### Weather expectations (late May — early June, Balkans)

| Elevation | Day temp | Night temp | Rain risk | Notes |
|---|---|---|---|---|
| < 500m | 20-30°C | 12-18°C | Moderate | Can be hot in valleys |
| 500-1000m | 15-25°C | 8-14°C | Moderate | |
| 1000-1500m | 10-20°C | 4-10°C | High | Mountain weather changes fast |
| > 1500m | 5-15°C | 0-6°C | High | Possible frost above 1800m |

## Mental strategy

### The five phases of ultra-racing

| Phase | When | What happens | What to do |
|---|---|---|---|
| **Euphoria** | Day 1 | Everything feels great, adrenaline high | Rein it in. You know the cost of going too hard on Day 1. |
| **Settling** | Days 2-3 | Finding rhythm, body adjusting | Establish routine. Eat, ride, sleep on schedule. |
| **The Pit** | Days 4-5 | Fatigue peaks, motivation drops, body hurts everywhere | **Keep riding.** This passes. Eat something warm. Call home. Small goals (next resupply). |
| **Acceptance** | Days 6-7 | Pain becomes background noise, robot mode | Autopilot is OK. Just keep the routine going. |
| **Finish smell** | Day 7+ | Energy returns, excitement builds | Don't take risks. No heroic descents. The finish is coming. |

### Decision rules

Pre-make these decisions so you don't have to think when exhausted:

1. **"Should I keep riding?"** → If it's before 00:30: yes. If after 02:00: stop immediately.
2. **"Should I eat?"** → Yes. Always yes. Even if not hungry.
3. **"Should I skip sleep?"** → No. Never after Day 2.
4. **"Should I push harder?"** → No. Not until Day 7 and the finish is within reach.
5. **"Should I quit?"** → Sleep on it. Literally. If you still want to quit after sleeping, re-evaluate. Most "I want to quit" feelings pass with food + sleep.

### Mantras

- *"I've done this before."* (You have. Hydra was 8 days.)
- *"Just ride to the next resupply."* (Never think about the finish. Think about the next 50 km.)
- *"Eat. Drink. Pedal. Repeat."*

## Hardest sections — know what's coming

```{r hardest-sections}
# Find 50km windows with most climbing
windows <- tibble(start_km = seq(0, max(segments$cum_dist) - 50, by = 25)) |>
  rowwise() |>
  mutate(
    end_km = start_km + 50,
    climb = sum(segments$gain_m[segments$cum_dist > start_km & segments$cum_dist <= end_km]),
    time_h = sum(segments$time_h[segments$cum_dist > start_km & segments$cum_dist <= end_km]),
    avg_speed = 50 / time_h,
    day = findInterval(riding_time_at_dist(start_km), day_info$cum_riding) + 1L
  ) |>
  ungroup() |>
  filter(climb > 1100) |>
  arrange(desc(climb))

windows |>
  head(10) |>
  mutate(Section = sprintf("km %.0f–%.0f", start_km, end_km),
         Climb = sprintf("%.0fm", climb),
         `Avg speed` = sprintf("%.1f km/h", avg_speed),
         `Time` = sprintf("%.1fh", time_h)) |>
  select(Section, Climb, `Avg speed`, Time, Day = day) |>
  kable() |> kable_styling(full_width = FALSE) |>
  row_spec(1:3, bold = TRUE)
```

## Summary: the plan in one page

::: {.callout-tip}
## Race in 7 words
**Start slow. Eat always. Sleep enough. Finish.**
:::

| Day | Date | Target km | Climbing | Key waypoints | Stop |
|---|---|---|---|---|---|
| 1 | Fri 29 May | ~179 | 4000m | Fužine (km 164) | 00:30, sleep 6h |
| 2 | Sat 30 May | ~230 | 3800m | Gospić (km 345) | 02:00, sleep 4.5h |
| 3 | Sun 31 May | ~237 | 3713m | **CP1 Mazin** (km 451), Knin (km 537) | 02:00, sleep 4.5h |
| 4 | Mon 1 Jun | ~236 | 3539m | Livno (km 664), **Mostar** (km 812) | 02:00, sleep 4.5h |
| 5 | Tue 2 Jun | ~209 | 4102m | CP2 approach, remote Bosnia | 00:30, sleep 6h |
| 6 | Wed 3 Jun | ~200 | **4619m** | **CP2 Kopilovi**, Žabljak (km 1160) | 00:30, sleep 6h |
| 7 | Thu 4 Jun | ~97 | 715m | Nikšić → **Risan FINISH** | **Done** |

**Realistic finish: Friday 5 June or Saturday 6 June** — well within the 9-day target.

<hr>
*Generated `r Sys.Date()` using GPX route analysis, Strava training data (FTP `r FTP`W), and Hydra Epic 2025 race data.*
