---
title: "Trans Balkan Race 2026 — Race Plan"
date: "2026-05-29"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    embed-resources: true
    fig-width: 12
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(xml2)
library(knitr)

# ── Parameters ────────────────────────────────────────────────────────────────

GPX_FILE        <- "Trans Balkan Race 2026 __ Issue 1.GPX"
SLEEP_HOUR      <- 2      # target sleep time (02:00)
SLEEP_DURATION  <- 4.5    # minimum sleep hours; wake target = SLEEP_HOUR + SLEEP_DURATION = 06:30
MAX_SLEEP       <- 6      # maximum sleep hours; earliest valid stop = 06:30 − 6h = 00:30
MIN_RIDING_H    <- 10     # minimum riding hours per day (safety net against bridge days)
MAX_DAILY_CLIMB <- 4000   # m elevation gain/day (soft: enforced only after earliest stop = 00:30)
BASE_SPEED      <- 13.8   # km/h on flat terrain
MAX_DESCENT     <- 32     # km/h cap on descents
UPHILL_K        <- 0.08   # exponential penalty per % grade uphill
DOWNHILL_K      <- 0.04   # exponential gain per % grade downhill

race_start <- as.POSIXct("2026-05-29 09:00:00", tz = "Europe/Paris")

# ── Parse GPX ─────────────────────────────────────────────────────────────────

gpx  <- read_xml(GPX_FILE)
ns   <- c(g = "http://www.topografix.com/GPX/1/1")
pts  <- xml_find_all(gpx, ".//g:trkpt", ns)

track <- tibble(
  lat = as.numeric(xml_attr(pts, "lat")),
  lon = as.numeric(xml_attr(pts, "lon")),
  ele = as.numeric(xml_text(xml_find_first(pts, "g:ele", ns)))
) |>
  mutate(ele = zoo::rollmedian(ele, k = 31, fill = NA, align = "center")) |>
  filter(!is.na(ele))

haversine <- function(lat1, lon1, lat2, lon2) {
  r    <- 6371
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  a    <- sin(((lat2-lat1)*pi/180)/2)^2 +
          cos(phi1)*cos(phi2)*sin(((lon2-lon1)*pi/180)/2)^2
  2 * r * asin(sqrt(a))
}

segments <- track |>
  mutate(
    dist_km  = haversine(lag(lat), lag(lon), lat, lon),
    ele_diff = ele - lag(ele),
    grade    = pmax(pmin(ifelse(dist_km > 0, (ele_diff/(dist_km*1000))*100, 0), 30), -30)
  ) |>
  slice(-1) |>
  filter(!is.na(dist_km), dist_km > 0) |>
  mutate(
    speed_kmh = case_when(
      grade >= 0 ~ BASE_SPEED * exp(-UPHILL_K * grade),
      grade <  0 ~ pmin(BASE_SPEED * exp(-DOWNHILL_K * grade), MAX_DESCENT)
    ),
    time_h   = dist_km / speed_kmh,
    gain_m   = pmax(ele_diff, 0),
    cum_dist = cumsum(dist_km),
    cum_time = cumsum(time_h),
    cum_gain = cumsum(gain_m)
  )

total_dist  <- sum(segments$dist_km)
total_climb <- sum(segments$gain_m)
total_time  <- sum(segments$time_h)

# ── Day simulation ─────────────────────────────────────────────────────────────
# Elevation cap is SOFT: only enforced after earliest valid stop (00:30).
# This eliminates short bridge days and keeps daily distances equal.

simulate_days <- function(segs, start_wall) {
  WAKE_HOUR <- SLEEP_HOUR + SLEEP_DURATION   # 6.5 = 06:30
  days_list <- list()
  seg_idx   <- 1L
  wall_time <- start_wall

  while (seg_idx <= nrow(segs)) {
    day_h0          <- as.numeric(difftime(wall_time, floor_date(wall_time, "day"), units = "hours"))
    hrs_to_latest   <- (SLEEP_HOUR            - day_h0 + 24) %% 24
    hrs_to_earliest <- (WAKE_HOUR - MAX_SLEEP - day_h0 + 24) %% 24
    if (hrs_to_latest   == 0) hrs_to_latest   <- 24
    if (hrs_to_earliest == 0) hrs_to_earliest <- 24

    day_riding <- 0; day_climb <- 0; day_dist <- 0; day_start <- wall_time

    while (seg_idx <= nrow(segs)) {
      s <- segs[seg_idx, ]
      if (day_riding + s$time_h > hrs_to_latest) break
      if (day_climb  + s$gain_m > MAX_DAILY_CLIMB &&
          day_riding >= MIN_RIDING_H              &&
          day_riding >= hrs_to_earliest) break
      day_riding <- day_riding + s$time_h
      day_dist   <- day_dist   + s$dist_km
      day_climb  <- day_climb  + s$gain_m
      seg_idx    <- seg_idx + 1L
    }

    stop_wall <- day_start + as.difftime(day_riding, units = "hours")
    next_0630 <- floor_date(stop_wall, "day") + as.difftime(WAKE_HOUR, units = "hours")
    if (next_0630 <= stop_wall) next_0630 <- next_0630 + days(1)
    sleep_h   <- max(SLEEP_DURATION, min(MAX_SLEEP,
                   as.numeric(difftime(next_0630, stop_wall, units = "hours"))))

    days_list[[length(days_list) + 1]] <- list(
      day_start_wall = day_start,
      day_end_wall   = stop_wall,
      riding_h       = day_riding,
      dist_km        = day_dist,
      climb_m        = day_climb,
      sleep_h        = sleep_h
    )
    wall_time <- stop_wall + as.difftime(sleep_h, units = "hours")
    if (seg_idx > nrow(segs)) break
  }

  bind_rows(lapply(days_list, as_tibble)) |>
    mutate(day = row_number(), cum_dist = cumsum(dist_km), cum_riding = cumsum(riding_h))
}

day_info <- simulate_days(segments, race_start)

total_days  <- nrow(day_info) - 1 +
  (total_time - day_info$cum_riding[nrow(day_info) - 1]) / day_info$riding_h[nrow(day_info)]
finish_wall <- day_info$day_start_wall[nrow(day_info)] +
  as.difftime(total_time - c(0, day_info$cum_riding)[nrow(day_info)], units = "hours")

# ── Helpers ────────────────────────────────────────────────────────────────────

riding_time_at_dist <- function(dist_km) {
  approx(segments$cum_dist, segments$cum_time, xout = dist_km, rule = 2)$y
}

wall_clock <- function(riding_time_h) {
  day_idx   <- pmin(findInterval(riding_time_h, day_info$cum_riding) + 1L, nrow(day_info))
  cum_start <- c(0, day_info$cum_riding)[day_idx]
  day_info$day_start_wall[day_idx] + as.difftime(riding_time_h - cum_start, units = "hours")
}

# ── Checkpoints ────────────────────────────────────────────────────────────────

checkpoints <- tribble(
  ~name,             ~cum_dist_km,
  "CP1 Mazin",              451,
  "CP2 Kopilovi",          1060,
  "Finish Risan",          1395
)

# ── Daily table data ───────────────────────────────────────────────────────────

daily_tbl <- day_info |>
  rowwise() |>
  mutate(
    start_dist = if (day == 1) 0 else day_info$cum_dist[day - 1],
    cps = paste(
      checkpoints$name[checkpoints$cum_dist_km > start_dist &
                       checkpoints$cum_dist_km <= cum_dist],
      collapse = ", "
    )
  ) |>
  ungroup() |>
  mutate(
    date_window  = sprintf("%s %s–%s",
                           format(day_start_wall, "%a %d %b"),
                           format(day_start_wall, "%H:%M"),
                           format(day_end_wall,   "%H:%M")),
    cum_dist_fmt = if_else(cum_dist >= total_dist,
                           sprintf("%.0f ✓", total_dist),
                           sprintf("%.0f",   cum_dist))
  )

# Truncate at finish row (simulation naturally stops at last segment)
finish_row <- which(daily_tbl$cum_dist >= total_dist - 0.01)[1]
if (is.na(finish_row)) finish_row <- nrow(daily_tbl)
daily_tbl  <- daily_tbl[1:finish_row, ]

# ── Resupply table data ────────────────────────────────────────────────────────

resupply <- tribble(
  ~location,              ~km,   ~type,         ~on_route,
  "Kozina (SLO)",         16.6,  "Village",      TRUE,
  "Slavnik (SLO)",        28.2,  "Mountain hut", TRUE,
  "Dane (CRO)",           47.4,  "Fountain",     TRUE,
  "Sveti Pavel (SLO)",    74.2,  "Fountain",     TRUE,
  "Platak (CRO)",        141,    "Mountain hut", TRUE,
  "Fužine (CRO)",        164,    "Village",      TRUE,
  "Oltari (CRO)",        245,    "Fountain",     TRUE,
  "Dom Zavižan (CRO)",   260,    "Mountain hut", TRUE,
  "Gospić (CRO)",        345,    "City",         TRUE,
  "CP1 Mazin (CRO)",     451,    "Checkpoint",   TRUE,
  "Velika Popina (CRO)", 490,    "Fountain",     TRUE,
  "Knin (CRO)",          537,    "City",         TRUE,
  "Livno (BiH)",         664,    "Village",      TRUE,
  "Šuica (BiH)",         701,    "Village",      TRUE,
  "Mostar (BiH)",        812,    "City",         TRUE,
  "Nevesinje (BiH)",     858,    "Village",      FALSE,
  "Ulog (BiH)",          897,    "Village",      TRUE,
  "Fountain (BiH)",      938,    "Fountain",     TRUE,
  "Miljevina (BiH)",     990,    "Village",      FALSE,
  "Popov Most (BiH)",   1034,    "Village",      TRUE,
  "Brod (BiH)",         1055,    "Village",      TRUE,
  "CP2 Kopilovi (BiH)", 1060,    "Checkpoint",   TRUE,
  "Žabljak (MNE)",      1160,    "City",         TRUE,
  "Njegovuđa (MNE)",    1170,    "Village",      TRUE,
  "Kolašin (MNE)",      1228,    "City",         FALSE,
  "Nikšić (MNE)",       1325,    "City",         TRUE,
  "Risan / Finish",     1395,    "Finish",       TRUE
) |>
  mutate(
    riding_h = map_dbl(km, riding_time_at_dist),
    day_num  = findInterval(riding_h, day_info$cum_riding) + 1L,
    arrival  = do.call(c, map(riding_h, wall_clock)),
    gap_km   = km - lag(km, default = 0)
  )
```

## Route overview

```{r stats}
tibble(
  Metric = c("Total distance", "Total climb", "Total descent",
             "Effective riding time", "Max daily climb",
             "Days", "Estimated finish"),
  Value  = c(
    sprintf("%.0f km",  total_dist),
    sprintf("%.0f m",   total_climb),
    sprintf("%.0f m",   sum(pmax(-segments$ele_diff, 0), na.rm = TRUE)),
    sprintf("%.1f h",   total_time),
    sprintf("%.0f m",   MAX_DAILY_CLIMB),
    sprintf("%.1f",     total_days),
    format(finish_wall, "%a %d %b %Y %H:%M CEST")
  )
) |> kable(col.names = c("", ""))
```

**Start:** Trieste, Italy — Fri 29 May 2026, 09:00 CEST
**Finish:** Risan, Kotor Bay, Montenegro
**Sleep:** 02:00 · `r SLEEP_DURATION`–`r MAX_SLEEP` h · wake 06:30
**Elevation cap:** `r MAX_DAILY_CLIMB` m gain/day, soft — only enforced after 00:30 (prevents bridge days)
**Speed model:** flat `r BASE_SPEED` km/h · uphill exp(−`r UPHILL_K` × grade%) · descent capped `r MAX_DESCENT` km/h

## Elevation profile

```{r elevation-profile, fig.height=4}
elev_profile <- segments |>
  select(cum_dist, ele, cum_time) |>
  mutate(day = factor(pmin(findInterval(cum_time, day_info$cum_riding) + 1L, nrow(day_info))))

day_lines <- day_info |>
  filter(cum_dist < total_dist) |>
  mutate(label = sprintf("Day %d\n%.0f km", day + 1, cum_dist))

ggplot(elev_profile, aes(cum_dist, ele)) +
  geom_area(aes(fill = day), alpha = 0.55) +
  geom_vline(data = day_lines, aes(xintercept = cum_dist),
             linetype = "dashed", colour = "grey40", linewidth = 0.35) +
  geom_text(data = day_lines,
            aes(x = cum_dist, y = max(elev_profile$ele) * 0.95, label = label),
            size = 2.8, hjust = 1.08, colour = "grey30") +
  geom_vline(data = checkpoints, aes(xintercept = cum_dist_km),
             linetype = "dotted", colour = "firebrick", linewidth = 0.5) +
  geom_text(data = checkpoints, aes(x = cum_dist_km, y = 80, label = name),
            size = 2.6, hjust = -0.05, colour = "firebrick") +
  scale_fill_brewer(palette = "Set2", name = "Day") +
  scale_x_continuous(breaks = seq(0, 1400, 100), minor_breaks = seq(0, 1400, 50)) +
  labs(x = "Distance from start (km)", y = "Elevation (m)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_line(linewidth = 0.2))
```

## Daily plan

```{r daily-table}
daily_tbl |>
  select(
    Day             = day,
    `Date / window` = date_window,
    `Ride h`        = riding_h,
    `Sleep h`       = sleep_h,
    `km`            = dist_km,
    `Climb m`       = climb_m,
    `Cum. km`       = cum_dist_fmt,
    Checkpoints     = cps
  ) |>
  mutate(
    `Ride h`  = sprintf("%.1f", `Ride h`),
    `Sleep h` = sprintf("%.1f", `Sleep h`),
    `km`      = sprintf("%.0f", `km`),
    `Climb m` = sprintf("%.0f", `Climb m`)
  ) |>
  kable(align = c("c","l","r","r","r","r","r","l"))
```

## Resupply schedule

Days end at 02:00 or when `r MAX_DAILY_CLIMB` m of gain is reached — but the cap is only enforced after 00:30 (earliest valid stop), so no short bridge days occur.
Sleep `r SLEEP_DURATION`–`r MAX_SLEEP` h, all days resume at 06:30.
Times marked \* are off-route stops.

```{r resupply-table}
resupply |>
  mutate(
    arrival_fmt = format(arrival, "%a %d %b %H:%M"),
    location    = if_else(!on_route, paste0(location, " *"), location),
    gap_km      = sprintf("%.0f", gap_km),
    km          = sprintf("%.0f", km)
  ) |>
  select(
    Location             = location,
    `km`                 = km,
    `+km`                = gap_km,
    Type                 = type,
    Day                  = day_num,
    `Est. arrival (CEST)` = arrival_fmt
  ) |>
  kable(align = c("l","r","r","l","c","l")) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::row_spec(
    which(resupply$type == "Checkpoint"),
    bold = TRUE, background = "#fff3cd"
  ) |>
  kableExtra::row_spec(
    nrow(resupply),
    bold = TRUE, background = "#d4edda"
  )
```

## Speed model

The estimated times use a gradient-adjusted speed model:

$$
v(g) = \begin{cases}
v_0 \cdot e^{-k_u \cdot g} & g \geq 0 \text{ (uphill)} \\
\min\!\left(v_0 \cdot e^{-k_d \cdot g},\, v_{\max}\right) & g < 0 \text{ (downhill)}
\end{cases}
$$

where $g$ is gradient in %, $v_0 =$ `r BASE_SPEED` km/h, $k_u =$ `r UPHILL_K`, $k_d =$ `r DOWNHILL_K`, $v_{\max} =$ `r MAX_DESCENT` km/h.

```{r speed-curve, fig.height=3, fig.width=6}
tibble(grade = seq(-20, 20, 0.1)) |>
  mutate(speed = case_when(
    grade >= 0 ~ BASE_SPEED * exp(-UPHILL_K * grade),
    grade <  0 ~ pmin(BASE_SPEED * exp(-DOWNHILL_K * grade), MAX_DESCENT)
  )) |>
  ggplot(aes(grade, speed)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_hline(yintercept = BASE_SPEED, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 0.4, y = BASE_SPEED + 0.8,
           label = sprintf("flat: %.0f km/h", BASE_SPEED),
           hjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(breaks = seq(-20, 20, 5), labels = function(x) paste0(x, "%")) +
  labs(x = "Gradient", y = "Speed (km/h)") +
  theme_minimal(base_size = 11)
```

GPS elevation noise suppressed with rolling median (window = 31 points ≈ 300 m).
Gradient capped at ±30% before speed calculation.
