---
title: "Trans Balkan Race 2026 — Race Plan"
date: "2026-05-29"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    embed-resources: true
    fig-width: 12
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("tbr_plan.R")
library(knitr)

# ── Build daily plan ──────────────────────────────────────────────────────────

track    <- load_track(GPX_FILE)
segments <- compute_segments(track)
day_info <- simulate_days(segments, RACE_START)

total_dist  <- sum(segments$dist_km)
total_climb <- sum(segments$gain_m)
total_time  <- sum(segments$time_h)
total_days  <- nrow(day_info) - 1 +
  (total_time - day_info$cum_riding[nrow(day_info) - 1]) / day_info$riding_h[nrow(day_info)]
finish_wall <- day_info$day_start_wall[nrow(day_info)] +
  as.difftime(total_time - c(0, day_info$cum_riding)[nrow(day_info)], units = "hours")

# ── Tables for the report ─────────────────────────────────────────────────────

daily_tbl <- day_info |>
  rowwise() |>
  mutate(
    start_dist = if (day == 1) 0 else day_info$cum_dist[day - 1],
    cps = paste(
      CHECKPOINTS$name[CHECKPOINTS$cum_dist_km > start_dist &
                       CHECKPOINTS$cum_dist_km <= cum_dist],
      collapse = ", "
    )
  ) |>
  ungroup() |>
  mutate(
    date_window  = sprintf("%s %s–%s",
                           format(day_start_wall, "%a %d %b"),
                           format(day_start_wall, "%H:%M"),
                           format(day_end_wall,   "%H:%M")),
    cum_dist_fmt = if_else(cum_dist >= total_dist,
                           sprintf("%.0f ✓", total_dist),
                           sprintf("%.0f",   cum_dist))
  )

# Truncate at finish row (simulation naturally stops at last segment)
finish_row <- which(daily_tbl$cum_dist >= total_dist - 0.01)[1]
if (is.na(finish_row)) finish_row <- nrow(daily_tbl)
daily_tbl  <- daily_tbl[1:finish_row, ]

resupply <- build_resupply(RESUPPLY, segments, day_info)
```

## Route overview

```{r stats}
tibble(
  Metric = c("Total distance", "Total climb", "Total descent",
             "Effective riding time", "Max daily climb",
             "Days", "Estimated finish"),
  Value  = c(
    sprintf("%.0f km",  total_dist),
    sprintf("%.0f m",   total_climb),
    sprintf("%.0f m",   sum(pmax(-segments$ele_diff, 0), na.rm = TRUE)),
    sprintf("%.1f h",   total_time),
    sprintf("%.0f m",   MAX_DAILY_CLIMB),
    sprintf("%.1f",     total_days),
    format(finish_wall, "%a %d %b %Y %H:%M CEST")
  )
) |> kable(col.names = c("", ""))
```

**Start:** Trieste, Italy — Fri 29 May 2026, 09:00 CEST
**Finish:** Risan, Kotor Bay, Montenegro
**Sleep:** 02:00 · `r SLEEP_DURATION`–`r MAX_SLEEP` h · wake 06:30
**Elevation cap:** `r MAX_DAILY_CLIMB` m gain/day, soft — only enforced after 00:30 (prevents bridge days)
**Speed model:** flat `r BASE_SPEED` km/h · uphill exp(−`r UPHILL_K` × grade%) · descent capped `r MAX_DESCENT` km/h

## Route map

```{r route-map}
library(leaflet)

# Thin track for performance (~every 10th point)
thin_idx   <- seq(1, nrow(track), by = 10)
track_thin <- track[thin_idx, ]

# Assign day to each thinned point via cumulative time
cum_dist_thin <- cumsum(c(0, haversine(
  head(track_thin$lat, -1), head(track_thin$lon, -1),
  tail(track_thin$lat, -1), tail(track_thin$lon, -1)
)))
cum_time_thin <- approx(segments$cum_dist, segments$cum_time,
                        xout = cum_dist_thin, rule = 2)$y
track_thin$day <- pmin(findInterval(cum_time_thin, day_info$cum_riding) + 1L, nrow(day_info))

day_colors <- RColorBrewer::brewer.pal(max(track_thin$day), "Set2")

# Resupply markers
resupply_icons <- awesomeIcons(
  icon = case_when(
    resupply$type == "Checkpoint" ~ "flag",
    resupply$type == "City"       ~ "home",
    resupply$type == "Finish"     ~ "flag-checkered",
    TRUE                          ~ "tint"
  ),
  markerColor = case_when(
    resupply$type == "Checkpoint" ~ "orange",
    resupply$type == "City"       ~ "blue",
    resupply$type == "Finish"     ~ "green",
    TRUE                          ~ "lightblue"
  ),
  library = "fa"
)

# Resupply lat/lon from nearest track point
resupply_coords <- map_dfr(resupply$km, function(d) {
  idx <- which.min(abs(segments$cum_dist - d))
  tibble(lat = segments$lat[idx], lon = segments$lon[idx])
})

m <- leaflet(width = "100%") |>
  addProviderTiles(providers$OpenStreetMap) |>
  addProviderTiles(providers$OpenTopoMap, group = "Topo") |>
  addLayersControl(baseGroups = c("OpenStreetMap", "Topo"),
                   options = layersControlOptions(collapsed = TRUE))

for (d in sort(unique(track_thin$day))) {
  pts <- track_thin[track_thin$day == d, ]
  m <- m |> addPolylines(lng = pts$lon, lat = pts$lat,
                         color = day_colors[d], weight = 3, opacity = 0.8,
                         group = paste("Day", d),
                         label = paste("Day", d))
}

m <- m |>
  addAwesomeMarkers(
    lng = resupply_coords$lon, lat = resupply_coords$lat,
    icon = resupply_icons,
    label = sprintf("%s (km %.0f, Day %d)", resupply$location, resupply$km, resupply$day_num),
    popup = sprintf("<b>%s</b><br>km %.0f · %s<br>Day %d · %s",
                    resupply$location, resupply$km, resupply$type,
                    resupply$day_num, format(resupply$arrival, "%a %d %b %H:%M"))
  )

m
```

## Elevation profile

```{r elevation-profile, fig.height=4}
elev_profile <- segments |>
  select(cum_dist, ele, cum_time) |>
  mutate(day = factor(pmin(findInterval(cum_time, day_info$cum_riding) + 1L, nrow(day_info))))

day_lines <- day_info |>
  filter(cum_dist < total_dist) |>
  mutate(label = sprintf("Day %d\n%.0f km", day + 1, cum_dist))

ggplot(elev_profile, aes(cum_dist, ele)) +
  geom_area(aes(fill = day), alpha = 0.55) +
  geom_vline(data = day_lines, aes(xintercept = cum_dist),
             linetype = "dashed", colour = "grey40", linewidth = 0.35) +
  geom_text(data = day_lines,
            aes(x = cum_dist, y = max(elev_profile$ele) * 0.95, label = label),
            size = 2.8, hjust = 1.08, colour = "grey30") +
  geom_vline(data = CHECKPOINTS, aes(xintercept = cum_dist_km),
             linetype = "dotted", colour = "firebrick", linewidth = 0.5) +
  geom_text(data = CHECKPOINTS, aes(x = cum_dist_km, y = 80, label = name),
            size = 2.6, hjust = -0.05, colour = "firebrick") +
  scale_fill_brewer(palette = "Set2", name = "Day") +
  scale_x_continuous(breaks = seq(0, 1400, 100), minor_breaks = seq(0, 1400, 50)) +
  labs(x = "Distance from start (km)", y = "Elevation (m)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_line(linewidth = 0.2))
```

## Daily plan

```{r daily-table}
daily_tbl |>
  select(
    Day             = day,
    `Date / window` = date_window,
    `Ride h`        = riding_h,
    `Sleep h`       = sleep_h,
    `km`            = dist_km,
    `Climb m`       = climb_m,
    `Cum. km`       = cum_dist_fmt,
    Checkpoints     = cps
  ) |>
  mutate(
    `Ride h`  = sprintf("%.1f", `Ride h`),
    `Sleep h` = sprintf("%.1f", `Sleep h`),
    `km`      = sprintf("%.0f", `km`),
    `Climb m` = sprintf("%.0f", `Climb m`)
  ) |>
  kable(align = c("c","l","r","r","r","r","r","l"))
```

## Resupply schedule

Days end at 02:00 or when `r MAX_DAILY_CLIMB` m of gain is reached — but the cap is only enforced after 00:30 (earliest valid stop), so no short bridge days occur.
Sleep `r SLEEP_DURATION`–`r MAX_SLEEP` h, all days resume at 06:30.
Times marked \* are off-route stops.

```{r resupply-table}
resupply |>
  mutate(
    arrival_fmt = format(arrival, "%a %d %b %H:%M"),
    location    = if_else(!on_route, paste0(location, " *"), location),
    gap_h_fmt   = sprintf("%.1f", gap_h),
    gap_km      = sprintf("%.0f", gap_km),
    km          = sprintf("%.0f", km)
  ) |>
  select(
    Location              = location,
    `km`                  = km,
    `+km`                 = gap_km,
    `+h`                  = gap_h_fmt,
    Type                  = type,
    Day                   = day_num,
    `Est. arrival (CEST)` = arrival_fmt
  ) |>
  kable(align = c("l","r","r","r","l","c","l")) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::row_spec(
    which(resupply$type == "Checkpoint"),
    bold = TRUE, background = "#fff3cd"
  ) |>
  kableExtra::row_spec(
    nrow(resupply),
    bold = TRUE, background = "#d4edda"
  )
```

## Speed model

The estimated times use a gradient-adjusted speed model:

$$
v(g) = \begin{cases}
v_0 \cdot e^{-k_u \cdot g} & g \geq 0 \text{ (uphill)} \\
\min\!\left(v_0 \cdot e^{-k_d \cdot g},\, v_{\max}\right) & g < 0 \text{ (downhill)}
\end{cases}
$$

where $g$ is gradient in %, $v_0 =$ `r BASE_SPEED` km/h, $k_u =$ `r UPHILL_K`, $k_d =$ `r DOWNHILL_K`, $v_{\max} =$ `r MAX_DESCENT` km/h.

```{r speed-curve, fig.height=3, fig.width=6}
tibble(grade = seq(-20, 20, 0.1)) |>
  mutate(speed = speed_for_grade(grade)) |>
  ggplot(aes(grade, speed)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_hline(yintercept = BASE_SPEED, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 0.4, y = BASE_SPEED + 0.8,
           label = sprintf("flat: %.0f km/h", BASE_SPEED),
           hjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(breaks = seq(-20, 20, 5), labels = function(x) paste0(x, "%")) +
  labs(x = "Gradient", y = "Speed (km/h)") +
  theme_minimal(base_size = 11)
```

GPS elevation noise suppressed with rolling median (window = 31 points ≈ 300 m).
Gradient capped at ±30% before speed calculation.
