---
title: "Trans Balkan Race 2026 — Race Plan"
date: "2026-05-29"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    embed-resources: true
    fig-width: 12
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(xml2)
library(knitr)

# ── Parameters ────────────────────────────────────────────────────────────────

GPX_FILE     <- "Trans Balkan Race 2026 __ Issue 1.GPX"
RIDING_HOURS <- 16      # riding hours per day
BASE_SPEED   <- 15      # km/h on flat terrain
MAX_DESCENT  <- 32      # km/h cap on descents
UPHILL_K     <- 0.08    # exponential speed penalty per % grade uphill
DOWNHILL_K   <- 0.04    # exponential speed gain per % grade downhill

race_start <- as.POSIXct("2026-05-29 09:00:00", tz = "Europe/Paris")

# ── Parse GPX ────────────────────────────────────────────────────────────────

gpx  <- read_xml(GPX_FILE)
ns   <- c(g = "http://www.topografix.com/GPX/1/1")
pts  <- xml_find_all(gpx, ".//g:trkpt", ns)

track <- tibble(
  lat = as.numeric(xml_attr(pts, "lat")),
  lon = as.numeric(xml_attr(pts, "lon")),
  ele = as.numeric(xml_text(xml_find_first(pts, "g:ele", ns)))
) |>
  mutate(ele = zoo::rollmedian(ele, k = 31, fill = NA, align = "center")) |>
  filter(!is.na(ele))

haversine <- function(lat1, lon1, lat2, lon2) {
  r    <- 6371
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dphi <- (lat2 - lat1) * pi / 180
  dlam <- (lon2 - lon1) * pi / 180
  a    <- sin(dphi/2)^2 + cos(phi1)*cos(phi2)*sin(dlam/2)^2
  2 * r * asin(sqrt(a))
}

segments <- track |>
  mutate(
    dist_km  = haversine(lag(lat), lag(lon), lat, lon),
    ele_diff = ele - lag(ele),
    grade    = pmax(pmin(ifelse(dist_km > 0, (ele_diff / (dist_km*1000))*100, 0), 30), -30)
  ) |>
  slice(-1) |>
  filter(!is.na(dist_km), dist_km > 0) |>
  mutate(
    speed_kmh = case_when(
      grade >= 0 ~ BASE_SPEED * exp(-UPHILL_K * grade),
      grade <  0 ~ pmin(BASE_SPEED * exp(-DOWNHILL_K * grade), MAX_DESCENT)
    ),
    time_h   = dist_km / speed_kmh,
    cum_dist = cumsum(dist_km),
    cum_time = cumsum(time_h)
  )

total_dist  <- sum(segments$dist_km)
total_climb <- sum(pmax(segments$ele_diff, 0), na.rm = TRUE)
total_time  <- sum(segments$time_h)
total_days  <- total_time / RIDING_HOURS

# ── Helpers ───────────────────────────────────────────────────────────────────

riding_time_at_dist <- function(dist_km) {
  approx(segments$cum_dist, segments$cum_time, xout = dist_km, rule = 2)$y
}

wall_clock <- function(riding_time_h) {
  day_num  <- pmax(ceiling(riding_time_h / RIDING_HOURS), 1)
  hours_in <- riding_time_h - (day_num - 1) * RIDING_HOURS
  race_start + days(day_num - 1) +
    hours(floor(hours_in)) + minutes(round((hours_in %% 1) * 60))
}

# ── Daily breakdown ───────────────────────────────────────────────────────────

checkpoints <- tribble(
  ~name,             ~cum_dist_km,
  "CP1 Mazin",              451,
  "CP2 Kopilovi",          1060,
  "Finish Risan",          1395
)

find_position <- function(time_limit) {
  i   <- max(which(segments$cum_time <= time_limit))
  rem <- time_limit - segments$cum_time[i]
  list(
    cum_dist = segments$cum_dist[i] + rem * segments$speed_kmh[min(i+1, nrow(segments))],
    cum_time = time_limit
  )
}

daily <- tibble(day = 1:ceiling(total_days), time_limit = day * RIDING_HOURS) |>
  mutate(
    pos      = map(time_limit, find_position),
    cum_dist = map_dbl(pos, "cum_dist"),
    date     = race_start + days(day - 1)
  ) |>
  select(-pos) |>
  mutate(day_dist = cum_dist - lag(cum_dist, default = 0))

# ── Resupply table ────────────────────────────────────────────────────────────

resupply <- tribble(
  ~location,              ~km,   ~type,         ~on_route,
  "Kozina (SLO)",         16.6,  "Village",      TRUE,
  "Slavnik (SLO)",        28.2,  "Mountain hut", TRUE,
  "Dane (CRO)",           47.4,  "Fountain",     TRUE,
  "Sveti Pavel (SLO)",    74.2,  "Fountain",     TRUE,
  "Platak (CRO)",        141,    "Mountain hut", TRUE,
  "Fužine (CRO)",        164,    "Village",      TRUE,
  "Oltari (CRO)",        245,    "Fountain",     TRUE,
  "Dom Zavižan (CRO)",   260,    "Mountain hut", TRUE,
  "Gospić (CRO)",        345,    "City",         TRUE,
  "CP1 Mazin (CRO)",     451,    "Checkpoint",   TRUE,
  "Velika Popina (CRO)", 490,    "Fountain",     TRUE,
  "Knin (CRO)",          537,    "City",         TRUE,
  "Livno (BiH)",         664,    "Village",      TRUE,
  "Šuica (BiH)",         701,    "Village",      TRUE,
  "Mostar (BiH)",        812,    "City",         TRUE,
  "Nevesinje (BiH)",     858,    "Village",      FALSE,
  "Ulog (BiH)",          897,    "Village",      TRUE,
  "Fountain (BiH)",      938,    "Fountain",     TRUE,
  "Miljevina (BiH)",     990,    "Village",      FALSE,
  "Popov Most (BiH)",   1034,    "Village",      TRUE,
  "Brod (BiH)",         1055,    "Village",      TRUE,
  "CP2 Kopilovi (BiH)", 1060,    "Checkpoint",   TRUE,
  "Žabljak (MNE)",      1160,    "City",         TRUE,
  "Njegovuđa (MNE)",    1170,    "Village",      TRUE,
  "Kolašin (MNE)",      1228,    "City",         FALSE,
  "Nikšić (MNE)",       1325,    "City",         TRUE,
  "Risan / Finish",     1395,    "Finish",       TRUE
) |>
  mutate(
    riding_h = map_dbl(km, riding_time_at_dist),
    day_num  = pmax(ceiling(riding_h / RIDING_HOURS), 1),
    arrival  = do.call(c, map(riding_h, wall_clock)),
    gap_km   = km - lag(km, default = 0)
  )
```

## Route overview

```{r stats}
tibble(
  Metric = c("Total distance", "Total climb", "Total descent",
             "Effective riding time", "Days at 16 h/day"),
  Value  = c(
    sprintf("%.0f km",  total_dist),
    sprintf("%.0f m",   total_climb),
    sprintf("%.0f m",   sum(pmax(-segments$ele_diff, 0), na.rm = TRUE)),
    sprintf("%.1f h",   total_time),
    sprintf("%.1f",     total_days)
  )
) |> kable(col.names = c("", ""))
```

**Start:** Trieste, Italy — Fri 29 May 2026, 09:00 CEST
**Finish:** Risan, Kotor Bay, Montenegro
**Model:** flat speed `r BASE_SPEED` km/h · uphill penalty `exp(−{r UPHILL_K} × grade%)` · descent capped at `r MAX_DESCENT` km/h

## Elevation profile

```{r elevation-profile, fig.height=4}
elev_profile <- segments |>
  mutate(day = factor(pmin(ceiling(cum_time / RIDING_HOURS), ceiling(total_days))))

day_lines <- daily |>
  filter(cum_dist < total_dist) |>
  mutate(label = sprintf("Day %d\n%.0f km", day + 1, cum_dist))

ggplot(elev_profile, aes(cum_dist, ele)) +
  geom_area(aes(fill = day), alpha = 0.55) +
  geom_vline(data = day_lines, aes(xintercept = cum_dist),
             linetype = "dashed", colour = "grey40", linewidth = 0.35) +
  geom_text(data = day_lines, aes(x = cum_dist, y = max(elev_profile$ele) * 0.95, label = label),
            size = 2.8, hjust = 1.08, colour = "grey30") +
  geom_vline(data = checkpoints, aes(xintercept = cum_dist_km),
             linetype = "dotted", colour = "firebrick", linewidth = 0.5) +
  geom_text(data = checkpoints, aes(x = cum_dist_km, y = 80, label = name),
            size = 2.6, hjust = -0.05, colour = "firebrick") +
  scale_fill_brewer(palette = "Set2", name = "Day") +
  scale_x_continuous(breaks = seq(0, 1400, 100), minor_breaks = seq(0, 1400, 50)) +
  labs(x = "Distance from start (km)", y = "Elevation (m)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_line(linewidth = 0.2))
```

## Daily plan

```{r daily-table}
daily_tbl <- daily |>
  mutate(
    day_dist_fmt = sprintf("%.0f km", day_dist),
    cum_dist_fmt = sprintf("%.0f km", pmin(cum_dist, total_dist)),
    date_fmt     = format(date, "%a %d %b")
  )

# Add checkpoints column
daily_tbl <- daily_tbl |>
  rowwise() |>
  mutate(
    start_dist = if (day == 1) 0 else daily$cum_dist[day - 1],
    cps = paste(
      checkpoints$name[checkpoints$cum_dist_km > start_dist &
                       checkpoints$cum_dist_km <= cum_dist],
      collapse = ", "
    )
  ) |>
  ungroup()

# Truncate at finish
finish_row <- which(daily_tbl$cum_dist >= total_dist)[1]
daily_tbl  <- daily_tbl[1:finish_row, ]
daily_tbl$cum_dist_fmt[finish_row] <- sprintf("%.0f km ✓", total_dist)

daily_tbl |>
  select(Day = day, Date = date_fmt,
         `km/day` = day_dist_fmt, `Cum. km` = cum_dist_fmt,
         Checkpoints = cps) |>
  kable(align = c("c","l","r","r","l"))
```

## Resupply schedule

Arrival times assume continuous riding of `r RIDING_HOURS` h/day starting at 09:00, with 8 h rest at the end of each riding day. Times marked \* are off-route stops.

```{r resupply-table}
resupply |>
  mutate(
    arrival_fmt = format(arrival, "%a %d %b %H:%M"),
    location    = if_else(!on_route, paste0(location, " *"), location),
    gap_km      = sprintf("%.0f", gap_km),
    km          = sprintf("%.0f", km)
  ) |>
  select(
    Location = location,
    `km` = km,
    `+km` = gap_km,
    Type = type,
    `Est. arrival (CEST)` = arrival_fmt
  ) |>
  kable(align = c("l","r","r","l","l")) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::row_spec(
    which(resupply$type == "Checkpoint"),
    bold = TRUE, background = "#fff3cd"
  ) |>
  kableExtra::row_spec(
    nrow(resupply),
    bold = TRUE, background = "#d4edda"
  )
```

## Speed model

The estimated times use a gradient-adjusted speed model:

$$
v(g) = \begin{cases}
v_0 \cdot e^{-k_u \cdot g} & g \geq 0 \text{ (uphill)} \\
\min\!\left(v_0 \cdot e^{-k_d \cdot g},\, v_{\max}\right) & g < 0 \text{ (downhill)}
\end{cases}
$$

where $g$ is gradient in %, $v_0 =$ `r BASE_SPEED` km/h, $k_u =$ `r UPHILL_K`, $k_d =$ `r DOWNHILL_K`, $v_{\max} =$ `r MAX_DESCENT` km/h.

```{r speed-curve, fig.height=3, fig.width=6}
tibble(grade = seq(-20, 20, 0.1)) |>
  mutate(speed = case_when(
    grade >= 0 ~ BASE_SPEED * exp(-UPHILL_K * grade),
    grade <  0 ~ pmin(BASE_SPEED * exp(-DOWNHILL_K * grade), MAX_DESCENT)
  )) |>
  ggplot(aes(grade, speed)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_hline(yintercept = BASE_SPEED, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 0.4, y = BASE_SPEED + 0.8,
           label = sprintf("flat: %.0f km/h", BASE_SPEED),
           hjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(breaks = seq(-20, 20, 5), labels = function(x) paste0(x, "%")) +
  labs(x = "Gradient", y = "Speed (km/h)") +
  theme_minimal(base_size = 11)
```

GPS elevation noise is suppressed with a rolling median (window = 31 points ≈ 300 m).
Gradient is capped at ±30% before speed calculation.
